P3064R1: How to Avoid OOTA Without Really Trying

### 2024-05-13 ootavacuous.tex
May 12 feedback from Alan Stern
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\paragraph{Invented Atomic Loads}
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Example Duplicated Load}
\label{lst:Example Duplicated Load}
\end{listing}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
extra cache misses for no good purpose.
And duplicating an atomic load can also lead to incorrect results if done
carelessly.
To see this, consider
Listing~\ref{lst:Example Duplicated Load}.
Suppose that the load on line~13 were duplicated, for example, by
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} probably
isn't prepared to deal with and that represents incorrect behavior,
that is, behavior that the abstract machine does not exhibit.
See
Appendix~\ref{app:Inventing Atomic Loads}
for more details on this example and other examples of invented
atomic loads.

@@

But if the compiler is permitted to invent atomic loads, then this
guarantee might not hold, as can be seen from this transformed version of
the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was 10, then the value of \co{r1} will be 21, which is not
a perfect square.

@@

Similar observations apply to atomic loads from pointers.
For example, consider
Listing~\ref{lst:Example Duplicated Load} on
page~\pageref{lst:Example Duplicated Load},
in which the abstract machine guarantees that line~17 passes two
fields from the same structure to \co{do_something_with()}.
This guarantee will not always be met if the compiler is permitted to
invent atomic loads such as that contained in the commented-out line~15.
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   do_something(fp);
15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   a = fp->a;
17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
18   b = fp->b;
19   do_something_with(a, b);
20 }
\end{verbatim}
\caption{Example Merging of Duplicated Loads}
\label{lst:Example Merging of Duplicated Loads}
\end{listing}
But suppose that there was an earlier load from \co{globalfoop}, as
shown on lines~13 and~14 in
Listing~\ref{lst:Example Merging of Duplicated Loads}.
Then, assuming that \co{do_something()} is synchronization-free, the
second load on line~15 could be removed without resulting in incorrect
behavior.

>>>>

\paragraph{Invented Atomic Loads}
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Example Duplicated Load}
\label{lst:Example Duplicated Load}
\end{listing}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
extra cache misses for no good purpose.
And duplicating an atomic load can also lead to incorrect results if done
carelessly.
To see this, consider
Listing~\ref{lst:Example Duplicated Load}.
Suppose that the load on line~13 were duplicated, for example, as if the
compiler had uncommented line~15.
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} probably
isn't prepared to deal with and that represents incorrect behavior,
that is, behavior that the abstract machine does not exhibit.
See
Appendix~\ref{app:Inventing Atomic Loads}
for more details on this example and other examples of invented
atomic loads.

@@

But if the compiler is permitted to invent atomic loads, then compiled
code might not respect this guarantee, thus (incorrectly) deviating
from the compiled code.
This issue is evident in this transformed version of the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was 10, then the value of \co{r1} will be 21, which is not
a perfect square.

@@

Similar observations apply to atomic loads from pointers.
For example, consider
Listing~\ref{lst:Example Duplicated Load} on
page~\pageref{lst:Example Duplicated Load},
in which the abstract machine guarantees that line~17 passes two
fields from the same structure to \co{do_something_with()}.
This guarantee will not always be met if the compiler is permitted to
invent atomic loads such as that contained in the commented-out line~15.
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   do_something(fp);
15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   a = fp->a;
17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
18   b = fp->b;
19   do_something_with(a, b);
20 }
\end{verbatim}
\caption{Example Merging of Duplicated Loads}
\label{lst:Example Merging of Duplicated Loads}
\end{listing}
But suppose that there was an earlier load from \co{globalfoop}, as
shown on lines~13 and~14 in
Listing~\ref{lst:Example Merging of Duplicated Loads}.
Then, assuming that \co{do_something()} is synchronization-free, the
second load on line~15 if that listing could be removed without resulting
in incorrect behavior.
